The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-17 write-up/covid-writeup.tex
Draft 2 done

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{hyperref}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage[margin=1.5in]{geometry}

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information to improve over simple random sampling, and the distinction between data quality and decision making in the current pandemic.

@@

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section \ref{section:testingcomments}.  As of May 8th, the US has performed $8,412,095$ tests in total.  Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.  The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.  After smoothing, the empirical prevalence is $9.1\%$ on May 8th.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.  Under these parameters, the effective sample size is $14$.  Recent proposals~\citep{Siddarth2020} have argued in favor of increasing overall testing capacity.  With more tests available, the relative sampling rate may reduce.  Supposing it drops to $1.2$ then the effective sample size will increase to $299$.  See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.   The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement error and the temporal nature of the pandemic.  We then discuss effective sampling methods and data versus decision-making.

@@

\section{Decision-making: data versus information.}

Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.  This conclusion confuses data and information.  Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.  There are whole fields dedicated to careful analysis in such settings (e.g., causal inference from observational studies).  Being a data skeptic, does not mean one must wait for high quality data to act, especially in high risk scenarios.  One aims to have high quality information.  Yet, in the current crisis, we have just that.  Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.  Social distancing, contact tracing and mass testing are all important.  Individuals and populations with low quality data and only basic information will make decisions that respect inherent risks in unknown unknowns. Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.  A full discussion of these topics is beyond the scope of the current article; we merely bring up the matter because we are well aware that discussions of data quality can sometimes be misinterpreted as claims that we do not have the requisite information to act immediately and decisively.  This is simply not the case.

\section{That which does not break us, makes us stronger (but potentially not smarter)}
\label{section:testingcomments}

A slightly altered version of Nietzsche's famous quote has become a mantra for post-pandemic thinking: \emph{that which which does not break us, makes us stronger}.  While a potentially true statement with respect to viral resistance, it is not clear that governments have yet to learn lessons on pandemic response. Governments are pushing increased testing capacity and contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing the disease from spreading.  We would argue that, after the current wave, understanding prevalence is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\cite{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary. Once we ``return to normal''\footnote{or at least the new normal}, it is clear that testing strategies should include a focus on early detection to help head off future outbreaks.

Governments implicitly argue that increased testing capacity will perform this task.  Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.  Random testing removes these effect modifiers, giving governments more information to fight the disease.  This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.

\section{Discussion}

There is nothing routine about COVID-19, including the corresponding statistical questions.  The goal of this paper was to point out questionable statistical routines.  Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.  The precision in the reported numbers gives the illusion of accuracy when what needs to be understood is the level of uncertainty. Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.

Not only that, in current crisis data quality is often tied to data quantity due to rationing system.  Increasing testing under test rationing lowers correlation.  There is a lot of hidden information in the selection indicator $I_j$.  Indeed, individuals must first seek testing and then call their primary health care provider must decide whether they met thresholds for testing.  Our analysis shows the limitations and caveats that come from analysis of this data in terms of estimation of prevalence, rates of change, and cross-country comparisons.  We show how one can build simple sensitivity analyses to understand the potential bias in case-count data using surveys that will become more prevalent in the next few months.

One alternative approach is model-based extensions of SIR models to try and capture both selection bias and measurement error issues; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.  Indeed, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp. 96]{CoxHink74}. In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.

>>>>

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{hyperref}
\usepackage{amsmath, amsthm, amsfonts}
% \usepackage[margin=1.5in]{geometry}

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.

@@

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section \ref{section:discussion}.  As of May 8th, the US has performed $8,412,095$ tests in total.  Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.  The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.  After smoothing, the empirical prevalence is $9.1\%$ on May 8th.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.  Under these parameters, the effective sample size is $14$.  Recent proposals~\citep{Siddarth2020} have argued in favor of increasing overall testing capacity.  With more tests available, the relative sampling rate may reduce.  Supposing it drops to $1.2$ then the effective sample size will increase to $299$.  See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.   The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement error and the temporal nature of the pandemic.  We then discuss effective sampling methods and their importance in the current pandemic.

@@

% \section{Decision-making: data versus information.}

% Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.  This conclusion confuses data and information.  Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.  There are whole fields dedicated to careful analysis in such settings (e.g., causal inference from observational studies).  Being a data skeptic, does not mean one must wait for high quality data to act, especially in high risk scenarios.  One aims to have high quality information.  Yet, in the current crisis, we have just that.  Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.  Social distancing, contact tracing and mass testing are all important.  Individuals and populations with low quality data and only basic information will make decisions that respect inherent risks in unknown unknowns. Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.  A full discussion of these topics is beyond the scope of the current article; we merely bring up the matter because we are well aware that discussions of data quality can sometimes be misinterpreted as claims that we do not have the requisite information to act immediately and decisively.  This is simply not the case.

\section{Discussion}
\label{section:discussion}

There is nothing routine about COVID-19, including the corresponding statistical questions.  The goal of this paper was to point out questionable statistical routines.  Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.  The precision in the reported numbers gives the illusion of accuracy when what needs to be understood is the level of uncertainty. Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.  We end this paper with a brief discussion of several important topics related to selection bias and measurement error.

\subsection*{Decision-making: data versus information}

Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.  This conclusion confuses data and information.  Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.  Being skeptical of observed data does not imply governments and communities should wait for high quality data to act; this is especially true in the current high risk scenario.  Instead, we should aim to have high quality information.  Yet in the current crisis we have just that.  Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.  Social distancing, contact tracing and mass testing are all important.  Individuals and communities with low quality data and only basic information will hopefully make decisions that respect inherent risks to their well being as well as the high degree of uncertainty in the observed data. Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.  A full discussion of these topics is beyond the scope of the current article.

\subsection*{High quality information on prevalence.}

Governments are pushing for increased testing capacity and robust contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.  We would argue that, after the current wave, understanding prevalence is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\cite{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.

\subsection*{Data quantity versus quality}

Governments implicitly argue that increased testing capacity will perform the above task.  Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates. Random testing removes these effect modifiers, giving governments more information to fight the disease.  This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.

% Not only that, in current crisis data quality is often tied to data quantity due to rationing system.  Increasing testing under test rationing lowers correlation.  There is a lot of hidden information in the selection indicator $I_j$.  Indeed, individuals must first seek testing and then call their primary health care provider must decide whether they met thresholds for testing.  Our analysis shows the limitations and caveats that come from analysis of this data in terms of estimation of prevalence, rates of change, and cross-country comparisons.  We show how one can build simple sensitivity analyses to understand the potential bias in case-count data using surveys that will become more prevalent in the next few months.

\subsection*{Model-based solutions.}

One alternative approach to what is presented in this article is a model-based extension of SIR models to try and capture both selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.  When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp. 96]{CoxHink74}. In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.
